\Title

### 2023-03-28 2023-03-28_HBP_summit.tex
Update on Overleaf.

\newcommand{\Acknowledgments}{AG and LP received funding from the European Union ERA-NET CHIST-ERA 2018 research and innovation program under grant agreement N° ANR-19-CHR3-0008-03 (``\href{http://aprovis3d.eu/}{APROVIS3D}''). HL and LP received funding from the ANR project  N° ANR-20-CE23-0021 (''\href{https://laurentperrinet.github.io/grant/anr-anr/}{AgileNeuroBot}''). %HL received funding from a University of Montreal Artificial Intelligence scholarship. For the purpose of open access, the author has applied a CC BY public copyright licence to any Author Accepted Manuscript version arising from this submission.

@@

\begin{center}
\color{blue}
\Large
\href{https://laurentperrinet.github.io/publication/grimaldi-22-icip/}{https://laurentperrinet.github.io
\\ /publication/grimaldi-22-icip/} \end{center}
\vspace{2cm}

\end{minipage}
%
\vspace*{-1.5cm}
\hrule
\vspace{5pt}
\hrule
%----------------------------------------------------------------------------------------
\begin{multicols}{3} % This is how many columns your poster will be broken into, a portrait poster is generally split into 2 columns
%----------------------------------------------------------------------------------------
%	ABSTRACT
%----------------------------------------------------------------------------------------
\section*{\color{\colorsec}Abstract \vspace{.3cm}  \hrule}
\vspace{-2.5cm}
\Abstract
%----------------------------------------------------------------------------------------
%	MLR
%----------------------------------------------------------------------------------------
\section*{\color{\colorsec}Heterogeneous delays model\hrule}
%
\vspace*{-2.5cm}
\subsection*{\color{\colorsubsec}Illustration}
%
\begin{figure}[H]
    \centering
    \includegraphics[width=0.980\linewidth]{figures/HDSNN_graph.pdf}
    \caption{\textbf{The core mechanism of the HD SNN model.} \textit{(Top-Left)}~Two spiking motifs are emitted from four presynaptic neurons. Once integrated by the synapses of the postsynaptic neuron~\textit{(Right)}, the spiking motifs are shifted in time by the synaptic delays and weighted by the synaptic weights~\textit{(Middle-Left)}. When they reach the soma of the postsynaptic neuron, the different spikes contribute to a modification of its membrane potential according to an activation function. In this example, we use the same activation function as for a Leaky Integrate and Fire neuron~\textit{(Bottom-Left)}. The first spiking motif is synchronized by the synaptic delays and causes a sudden rise in the membrane potential of the postsynaptic neuron. An output spike is emitted when the membrane potential reaches the threshold and it is then reset.   \textit{(Right)}~An illustration of a spiking neuron with different synaptic weights (represented by the thickness of the synapses) and different synaptic delays (represented by the length of the synapses). }%
    \label{fig:izhikevich}%
\end{figure}
%TODO: changer le temps d'occurence des spikes
%\vspace*{1.5cm}
\subsection*{\color{\colorsubsec}Mathematical formalism}
%
\textbf{Event stream:}
%The spikes reaching our SNN can be formalized as:
$$
\mybox{$\event = \{(\presynaddr_\arank, 1, \timev_\arank)\}_{\arank \in [1,\numevent]}$}
$$
where $\numevent \in \mathbb{N} $ is the total number of events, $\timev_\arank$ is the time occurrence of event number $\arank$ and $\presynaddr_\arank$ an associated address , which is typically in the form $\presynaddr_\arank=(x_\arank, y_\arank, \polev_\arank)$. \\
%
\\
\textbf{Spiking neuron with hetero-synaptic delays:}
%
$$
\mybox{$\synapse = \{(\presynaddr_\ranksyn, \postsynaddr_\ranksyn, \synapticweight_\ranksyn, \synapticdelay_\ranksyn)\}_{\ranksyn \in [1,\Nsyn]}$}
$$
is a set of $\Nsyn$ synapses, where each synapse $\synapse_\ranksyn$ is associated to a presynaptic address $\presynaddr_\ranksyn$, a postsynaptic address $\postsynaddr_\ranksyn$, a weight $\synapticweight_\ranksyn$ and a delay $\synapticdelay_\ranksyn$. We can define the synapses of the receptive field of neuron $\postsynaddr$ as such: $\synapse^\postsynaddr =  \{(\presynaddr_\ranksyn, \postsynaddr_\ranksyn, \synapticweight_\ranksyn, \synapticdelay_\ranksyn) \| \postsynaddr_\ranksyn=\postsynaddr\}_{\ranksyn \in [1,\Nsyn]}  \subset \synapse$.
They transform the event stream as input into: 
$$
\mybox{$\event_\postsynaddr = \{(\presynaddr_\arank, \synapticweight_\arank, \timev_\arank+\synapticdelay_\ranksyn) \| \presynaddr_\arank = \presynaddr_\ranksyn \}_{\arank \in [1,\numevent], \ranksyn \in \synapse^\postsynaddr}$}
$$

@@

\noindent By discretizing time, we can transform any event-based input from an event-based camera into a Boolean matrix $A \in \{0, 1 \}^{\Npol \times \Ntime \times \Nx \times \Ny}$ where the ones define the events. Using this dense representation, the processing of $A$ by neuron $\postsynaddr$ can be written as: %
%
$$
\mybox{$\forall \timev, \; \;
\mathcal{C}_\postsynaddr(x_\postsynaddr, y_\postsynaddr, \timev)
= \sum_{p, \delta_x, \delta_y, \synapticdelay_\timev} \kernel_\postsynaddr(p, \delta_x, \delta_y, \synapticdelay_\timev) \cdot A(p, x - \delta_x, y - \delta_y, \timev-\synapticdelay_\timev)
$}
$$
%
where $\kernel_\postsynaddr$ gives the weights as a function of relative addresses, $\delta_x$ and $\delta_y$, and synaptic delays $\delta_\timev$.\\
%
\\
\textbf{Multinomial logistic regression:}

To take advantage of the position invariance observed in images and exploited in convolutional neural networks, we can further assume that synaptic motifs should be similar across different positions, so we can define a spatio-temporal convolutional operator: $\mathcal{C}_\class = \kernel_\class \ast A$, where $\class$ combines different postsynaptic neurons sharing the same class. Then, the activation function of our model is a softmax function implementing a form of Multinomial Logistic Regression, in analogy to a spiking Winner-Takes-All network~\cite{nessler_bayesian_2013}:

@@

    \textbf{Applying HD-SNN to the task of motion detection.} \textit{(Left)}~We plot here as a raster plot a 2D representation of the input event stream (showing ON spikes in red and OFF spikes in blue for each presynaptic address and time). A spatio-temporal convolution is applied to the dense representation of the input with $2$ distinct convolutional kernels (the green and orange kernels) that will define the output channels. The convolution is summed over the two polarities. If you have two axes $X$ and $Y$ to represent the presynaptic addresses as for the pixel grid of a DVS, you obtain a 3D convolution. We restrict the illustration to a 2D representation and to 2 possible classes (green and orange) that are associated with different motion directions. \textit{(Middle)}~For each position (address, time), one can compute the activation resulting from the convolution. The output of the convolution is processed by the nonlinearity of the MLR model (i.e., the softmax function). The output of the MLR gives a probability for each class associated with a particular kernel (colored bars in the highlighted pixel). \textit{(Right)}~By adding a spiking mechanism, here a winner-takes-all associated to a thresholding, we obtain as output of the HD-SNN model a new spike train with the different spikes associated to a specific motion class. Note that the position of the output spikes does not systematically correspond to the position of the input spikes but only when enough evidence is obtained.}

@@

To test our model, we will quantify its ability to categorize different motions. In that order, we will first define a set of synthetic stimuli, natural images with a rigid motion or also \textit{Motion Clouds}~\cite{leon2012motion}, which are natural-like random textures for which we can control for velocity, among other parameters.

@@

        {\bf - Motion Detection Task.} \textit{(Left)}~We use large natural images ($256\times256$) in which an aperture ($64\times64$) extracts a cropped image around the view axis. To mimic the effect of a saccadic eye movement, the view axis moves according to a stepwise random walk. We show such a trajectory with a length of $128$ time steps (green line). \textit{(Right)}~Snapshots of the synthetic event stream at different time steps (start and end frames marked by a white and black dot, respectively). The dynamics of the cropped image translated according to the trajectory as a function of time produces a naturalistic movie, which is then transformed into an event-based representation. Mimicking the retina, this representation encodes proportional increases or decreases in luminance, i.e. ON (red) and OFF (blue) events, in each pixel of the image. In the lower left corner of the snapshots, the translation vector is shown in red as one of the possible classes of motion. Note the change in the direction of motion between the third and fourth image, and also that, due to the aperture problem, contours parallel to the motion emit relatively fewer spikes.}

@@

@@

\subsection*{\color{\colorsubsec}Learning hetero-synaptic delays}

@@

    {\centering

@@

    \includegraphics[width=\linewidth]{figures/motion_kernels.pdf}

@@

    	Representation of the weights for $8$ directions for one speed (among the $8 \times 3$ different kernels of the model) as learned on the dataset of naturalistic scenes. Directions are shown as red arrows on the left. The spatio-temporal kernels are represented as slices of spatial weights at different delays. Delays vary along the horizontal axis from the far right (delay of one step) to the left (delay of $12$ steps). Each image corresponds to the weights at a given delay, with excitatory and inhibitory weights in warm and cold colors, respectively. Due to the observed symmetry between the ON and OFF event streams, we report that the kernels for the OFF polarities are very similar and are not shown here.	Different kernels are selective to the different motion directions and we observe for all kernels an orientation preference perpendicular to the orientation. 

@@

    \includegraphics[width=0.99\linewidth]{figures/motion_clouds.pdf} %
    \caption{{\bf Role of stimulus parameters in the motion detection accuracy.} Accuracy as a function of (from left to right) the mean spatial frequency, the bandwidth in spatial frequency (from a grating-like (left) to isotropic textures (right)), the bandwidth in orientation (from isotropic textures (left) to grating-like (right)), the bandwidth in speed (from a rigid motion (left) to independent frames (right)). Note that these accuracies are computed both in the case where orientation of the synthetic texture is necessarily perpendicular to the motion ('no aperture' condition) or in the generic case where orientation is independent of direction ('aperture').}
    \label{fig:motion_clouds}

@@

\section*{\color{\colorsec}Conclusion \vspace{.2cm} \hrule}

@@

We have introduced a generic SNN using hetero-synaptic delays and shown how it compares favorably with a state-of-the-art event-based algorithm used for classification~\cite{lagorce_hots_2017}. This shows that we may use the precise timing of a spike to enhance neural computations.

@@

One advantage of our model is the generality of the approach.
Indeed, this supervised learning scheme can be extended to a novel task by defining a new set of supervision pairs (for instance supervised by local orientation) which would lead to the emergence of new kernels adapted to this new task.
This constitutes a major advantage over other algorithms which derive event-based algorithms from specific physical rules. %(see for instance~\cite{Benosman12} for computing the optic flow using the luminance conservation rule).
We aim at extending the application of this model on more generic datasets acquired in natural conditions for progressively more complex tasks such as time-to-contact maps, but also to neurophysiological data.

@@

@@

>>>>

\newcommand{\Acknowledgments}{AG and LP received funding from the European Union ERA-NET CHIST-ERA 2018 research and innovation program under grant agreement N° ANR-19-CHR3-0008-03 (``\href{http://aprovis3d.eu/}{APROVIS3D}''). LP received funding from the ANR project  N° ANR-20-CE23-0021 (''\href{https://laurentperrinet.github.io/grant/anr-anr/}{AgileNeuroBot}''). %HL received funding from a University of Montreal Artificial Intelligence scholarship. For the purpose of open access, the author has applied a CC BY public copyright licence to any Author Accepted Manuscript version arising from this submission.

@@

\begin{center}
\color{blue}
\Large
\href{https://laurentperrinet.github.io/publication/grimaldi-22-bc/}{https://laurentperrinet.github.io
\\ /publication/grimaldi-22-bc/} \end{center}
\vspace{2cm}

\end{minipage}
%
\vspace*{-1.5cm}
\hrule
\vspace{5pt}
\hrule
%----------------------------------------------------------------------------------------
\begin{multicols}{3} % This is how many columns your poster will be broken into, a portrait poster is generally split into 2 columns
%----------------------------------------------------------------------------------------
%	ABSTRACT
%----------------------------------------------------------------------------------------
\section*{\color{\colorsec}Abstract \vspace{.3cm}  \hrule}
\vspace{-2.5cm}
\Abstract
%----------------------------------------------------------------------------------------
%	MLR
%----------------------------------------------------------------------------------------
\section*{\color{\colorsec}Heterogeneous delays model (HD-SNN)\hrule}
%
\vspace*{-2.5cm}
\subsection*{\color{\colorsubsec}Illustration}
%
\begin{figure}[H]
    \centering
    \includegraphics[width=0.980\linewidth]{figures/HDSNN_graph.pdf}
    \caption{\textbf{The core mechanism of the HD-SNN model.} \textit{(Top-Left)}~Two spiking motifs are emitted from four presynaptic neurons. Once integrated by the synapses of the postsynaptic neuron~\textit{(Right)}, the spiking motifs are shifted in time by the synaptic delays and weighted by the synaptic weights~\textit{(Middle-Left)}. When they reach the soma of the postsynaptic neuron, the different spikes contribute to a modification of its membrane potential according to an activation function. In this example, we use the same activation function as for a Leaky Integrate and Fire neuron~\textit{(Bottom-Left)}. The first spiking motif is synchronized by the synaptic delays and causes a sudden rise in the membrane potential of the postsynaptic neuron. An output spike is emitted when the membrane potential reaches the threshold and it is then reset.   \textit{(Right)}~An illustration of a spiking neuron with different synaptic weights (represented by the thickness of the synapses) and different synaptic delays (represented by the length of the synapses). }%
    \label{fig:izhikevich}%
\end{figure}
%TODO: changer le temps d'occurence des spikes
%\vspace*{1.5cm}
\subsection*{\color{\colorsubsec}Mathematical formalism}
%
\textbf{Event stream:}
%The spikes reaching our SNN can be formalized as:
$$
\mybox{$\event = \{(\presynaddr_\arank, 1, \timev_\arank)\}_{\arank \in [1,\numevent]}$}
$$
where $\numevent \in \mathbb{N} $ is the total number of events, $\timev_\arank$ is the time occurrence of event number $\arank$ and $\presynaddr_\arank$ an associated address , which is typically in the form $\presynaddr_\arank=(x_\arank, y_\arank, \polev_\arank)$. \\
%
\\
\textbf{Spiking neuron with hetero-synaptic delays:}
%
$$
\mybox{$\synapse = \{(\presynaddr_\ranksyn, \postsynaddr_\ranksyn, \synapticweight_\ranksyn, \synapticdelay_\ranksyn)\}_{\ranksyn \in [1,\Nsyn]}$}
$$
is a set of $\Nsyn$ synapses, where each synapse $\synapse_\ranksyn$ is associated to a presynaptic address $\presynaddr_\ranksyn$, a postsynaptic address $\postsynaddr_\ranksyn$, a weight $\synapticweight_\ranksyn$ and a delay $\synapticdelay_\ranksyn$. We can define the synapses of the receptive field of neuron $\postsynaddr$ as such: $\synapse^\postsynaddr =  \{(\presynaddr_\ranksyn, \postsynaddr_\ranksyn, \synapticweight_\ranksyn, \synapticdelay_\ranksyn) \| \postsynaddr_\ranksyn=\postsynaddr\}_{\ranksyn \in [1,\Nsyn]}  \subset \synapse$.
They transform the event stream as input into: 
$$
\mybox{$\event_\postsynaddr = \{(\presynaddr_\arank, \synapticweight_\arank, \timev_\arank+\synapticdelay_\ranksyn) \| \presynaddr_\arank = \presynaddr_\ranksyn \}_{\arank \in [1,\numevent], \ranksyn \in \synapse^\postsynaddr}$}
$$

@@

\noindent By discretizing time, we can transform any event-based input from an event-based camera into a Boolean matrix $A \in \{0, 1 \}^{\Npol \times \Ntime \times \Nx \times \Ny}$ where the ones define the events. Using this dense representation, the processing of $A$ by neuron $\postsynaddr$ can be written as: %
%
$$
\mybox{$\forall \timev, \; \;
\mathcal{C}_\postsynaddr(x_\postsynaddr, y_\postsynaddr, \timev)
= \sum_{p, \delta_x, \delta_y, \synapticdelay_\timev} \kernel_\postsynaddr(p, \delta_x, \delta_y, \synapticdelay_\timev) \cdot A(p, x - \delta_x, y - \delta_y, \timev-\synapticdelay_\timev)
$}
$$
%
where $\kernel_\postsynaddr$ gives the weights as a function of relative addresses, $\delta_x$ and $\delta_y$, and synaptic delays $\delta_\timev$.\\
%
\\
\textbf{Multinomial logistic regression (MLR):}

\noindent We can further assume that synaptic motifs should be similar across different positions, so we can define a spatio-temporal convolutional operator: $\mathcal{C}_\class = \kernel_\class \ast A$, where $\class$ combines different postsynaptic neurons sharing the same class. Then, the activation function of our model is a softmax function implementing a form of MLR, in analogy to a spiking Winner-Takes-All network~\cite{nessler_bayesian_2013}:

@@

    \textbf{Applying HD-SNN to the motion detection task.} \textit{(Left)}~We plot here a 2D representation of the input event stream. A spatio-temporal convolution is applied to the dense representation of the input with $2$ distinct convolutional kernels (the green and orange kernels) that will define the output channels (if you have two axes $X$ and $Y$ to represent the presynaptic addresses as for the pixel grid of a DVS, you obtain a 3D convolution). \textit{(Middle)}~The output of the convolution is processed by the nonlinearity of the MLR model (i.e., the softmax function). \textit{(Right)}~By adding a spiking mechanism, we obtain a new spike train with the spikes associated to the different motion classes.}

@@

@@

        {\bf - Motion Detection Task.} \textit{(Left)}~We use large natural images ($256\times256$) in which an aperture ($64\times64$) extracts a cropped image around the view axis. To mimic the effect of a saccadic eye movement, the view axis moves according to a stepwise random walk. We show such a trajectory with a length of $128$ time steps (green line). \textit{(Right)}~Snapshots of the synthetic event stream at different time steps. The dynamics of the cropped image translated according to the trajectory as a function of time produces a naturalistic movie, which is then transformed into an event-based representation. Mimicking the retina, this representation encodes proportional increases or decreases in luminance, i.e. ON (red) and OFF (blue) events, in each pixel of the image. In the lower left corner of the snapshots, the translation vector is shown in red as one of the possible classes of motion.}

@@

%\vspace{1cm}
\columnbreak

@@

\subsection*{\color{\colorsubsec}Learning heterogeneous delays}

@@

@@

    \includegraphics[width=.8\linewidth]{figures/motion_kernels.pdf}

@@

    	\textbf{Representation of the weights} for $8$ directions for one speed (among the $8 \times 3$ different kernels of the model) as learned on the dataset of naturalistic scenes. Directions are shown as red arrows on the left. The spatio-temporal kernels are represented as slices of spatial weights at different delays. Due to the observed symmetry between the ON and OFF event streams, we report that the kernels for the OFF polarities are very similar and are not shown here. Different kernels are selective to the different motion directions and we observe for all kernels an orientation preference perpendicular to the orientation. 

@@

    \includegraphics[width=0.6\linewidth]{figures/quant_accuracy.pdf}
    \caption{\textbf{Accuracy as a function of computational load} for the HD-SNN model (blue dots) with error bars indicating the 5\% - 95\% quantiles and a sigmoid fit (blue line). The relative computational load (on a logarithmic axis) is controlled by changing the percentage of nonzero weights relative to the dense convolution kernel. As we prune the coefficients, we observe a stable accuracy value, with a drop observed at about 25 times fewer computations.}
    \label{fig:accuracy}

@@

%\section*{\color{\colorsec}Conclusion \vspace{.2cm} \hrule}
%\vspace{-2.5cm}

@@

%We have introduced a generic SNN using hetero-synaptic delays and shown how it compares favorably with a state-of-the-art event-based algorithm used for classification~\cite{lagorce_hots_2017}. This shows that we may use the precise timing of a spike to enhance neural computations.

@@

%One advantage of our model is the generality of the approach.
%Indeed, this supervised learning scheme can be extended to a novel task by defining a new set of supervision pairs (for instance supervised by local orientation) which would lead to the emergence of new kernels adapted to this new task.
%This constitutes a major advantage over other algorithms which derive event-based algorithms from specific physical rules. %(see for instance~\cite{Benosman12} for computing the optic flow using the luminance conservation rule).
%We aim at extending the application of this model on more generic datasets acquired in natural conditions for progressively more complex tasks such as time-to-contact maps, but also to neurophysiological data.

@@

%\vspace{-1cm}

@@

%\hrule
%\vspace{-1cm}
